---
title: "Oddstream ts/anomolies "
subtitle: "On aggregated FIFA data (all quantitative)."
author: "Nicholas Spyrison"
date: "11 Mar 2022"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r opts_chunk, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Gist

<!-- Before working with Klaus -->
I am going to brush up on working with time series (ts), de-mean/trend/season/period, As well as ts features. My data will be quite noisy, I want to see what is out there and work start brainstorming some ideas.


# Summary

In __oddstream__ we are looking at univariate time series and looking for anomalous behavior. There are 640 time series, for each of 1459 evenly spaced time samples.

```{r}
library(oddstream)
library(tidyverse)
t <- nrow(anomalous_stream)
f <- ncol(anomalous_stream)

data <- as_tibble(anomalous_stream) %>%
  gather() %>%
  mutate(key = rep((1:f), each = t), Time = rep(1:t, f)) %>%
  setNames(c("Cable", "Value", "Time"))
skimr::skim(data)

p <- data %>%
  ggplot(aes(x = Time, y = Cable, fill = Value)) +
  geom_tile() +
  scale_fill_gradientn(colours = c("#F0E442", "#000000", "#000000"), values = c(0, .1, max(anomalous_stream))) +
  scale_x_continuous(breaks = seq(0, 1400, 200)) +
  scale_y_continuous(breaks = seq(0, 600, 100)) +
  labs(x= "Time", y= "Time Series ID")
p
```


# Feature spaces (PC1:2)

14 time series features are extracted. PCA is used to approximate and visualize this.

```{r}
## 100 time intervals, 100 time series, extract 14 features
features <- extract_tsfeatures(anomalous_stream[1:100, 1:100]) 
## PCA on 14 features
pc       <- get_pc_space(features)
p2       <- gg_featurespace(pc)
p2
```

Can add density overlay as a visual guide to see distant points.

```{r}
# Perform a 2D kernel density estimation using MASS::kde2d() and display the results with contours.
p2 + ggplot2::geom_density_2d()
```


# Train vs test thinking

`find_odd_streams` takes sliding window approach to test if the features are 

```{r}
# Considers the first window  of the data set as the training set and the remaining as the test stream
train_data  <- anomalous_stream[1:100, ]
test_stream <- anomalous_stream[101:700, ]
output      <- find_odd_streams(train_data, test_stream, trials = 50)

dim(train_data)  ## first 100 time intervals for all TS
dim(test_stream) ## Remaining 1355 time intervals for all TS.
str(output)      ## $out_marix (sic), feature by TS 
```

Let's look 

```{r}
sum(output$out_marix) ## Count of flagged feats*TS deemed anomalous
hist(apply(output$out_marix, 1, sum),
     main = "# Anomalous intervals within 14 features",
     xlab = "# Anomalous intervals in 1 feature") ## Anomalous intervals by feat.
hist(apply(output$out_marix, 2, sum),
     main = "# Anomalous intervals within 640 TS",
     xlab = "# Anomalous intervals in 1 TS") ## Anomalous intervals by feat.
```