---
title: "Forecasting principles and practice"
subtitle: "Brainstorming example for noisy busy univariate TS"
author: "Nicholas Spyrison"
date: "12 Mar 2022"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r opts_chunk, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Gist

I am using this to vet and flesh out some idea for a noisy univariate time series. Typically, I think about time series features approaches. For example, Rob and Priyanga did several pieces with this approach for water quality data in streams. I will be referring to Rob & George's fpp2, https://otexts.com/fpp2/, for examples and inspiration as well. I will be focusing on a particularly noisy time series, while working in parallel with the content in the book. 


# Setup

We'll use melb-syd flight data as an example ts with noise and oddities.

```{r}
library(fpp2)
#library(ggplot2)

dat <- fpp2::melsyd[, "Economy.Class"]
str(dat)
sum(is.na(dat))
autoplot(dat) +
  ggtitle("Economy class passengers: Melbourne-Sydney") +
  xlab("Year") +
  ylab("Passengers ['000]")
```


# Autocorellation

The data set is quite mess and there are few small seasonal changes. I want to see if any autocorelation is picked up on as an indication of signal shining through the noise.

```{r}
ggAcf(dat, lag = 52 * (1993 - 1987) / 2)
```

Good, plenty of signal to account for. There seems to be near linear negative trend with possibly a few seasons/cycles. Below we use a few other fast higher-level visualization. They are not particularly insightful here, but can be ussuful in other more regular time series.

```{r}
ggseasonplot(dat)
ggsubseriesplot(dat)
gglagplot(dat)
```

# Box-cox transformation

We can use a box-cox transformation to distribute the data normally. The tranfsormation has one parameter, $\lambda$. A good value for this can be estimated with `BoxCox.lambda`. The text-book example for box-cox transformation is for growing/shrinking seasonal variation such as in the data set `elec`

```{r}
(lambda <- BoxCox.lambda(dat))
dat_bc <- BoxCox(dat,lambda)
## Normalized difference doesn't plot, cause of NA?
#dat_diff <- scale(dat) - scale(dat_bc)
require(patchwork)
autoplot(dat) +
  ylab("Passengers ['000]") +
  autoplot(dat_bc) +
  ylab(paste0("Passengers (BoxCox(", round(lambda), "))"))
```

Working through other examples I want to box-cox works best when: i) seasonal variation is sizable compared to noise and ii) seasonal variation changes through out the period. 


# Random walk forecasts

We can do some preliminary with random walk forecasting.

```{r}
fc  <- rwf(eggs, drift=TRUE, lambda=0, h=50, level=80)
fc2 <- rwf(eggs, drift=TRUE, lambda=0, h=50, level=80, biasadj=TRUE)
autoplot(eggs) +
  autolayer(fc,  series="Simple back transformation") +
  autolayer(fc2, series="Bias adjusted", PI=FALSE) +
  guides(colour=guide_legend(title="Forecast"))

fc <- rwf(dat, drift=FALSE, lambda=lambda, h=50, level=50)
fc_adj <- rwf(dat, drift=FALSE, lambda=lambda, h=50, level=50, biasadj=TRUE)
autoplot(dat) +
  autolayer(fc, series="Simple back transformation") +
  autolayer(fc_adj, series="Bias adjusted", PI=FALSE) +
  guides(colour=guide_legend(title="Forecast"))
```

# model summary with `checkresiduals()`

```{r}
checkresiduals(fc_adj)
```

Want symmetric white noise on top. If visuals patterns, drift, or seasonality is present this information is not captured by the model. Residuals ideally are normal; this is pretty good, few low values in the end of 1988. peaks and regular occilations in the ACF, indicate autocorrelation. This is corroborated with the Ljung-Box test (p = 9e-7), with a null hypothesis of independence. That is there is strong evidence suggesting that there auto-dependance in the model (that could presumably suggests a better model could be captured). 

# Piecewise and various forecasting methods

```{r}
dat
h <- 13
fit.lin    <- tslm(dat ~ trend + season)
fcasts.lin <- forecast(fit.lin, h = h)
fit.exp    <- tslm(dat ~ trend + season, lambda = BoxCox.lambda(dat))
fcasts.exp <- forecast(fit.exp, h = h)

t <- time(dat)
# t.break1 <- c(1989, 13) ## Didn't work
# t.break2 <- c(1990, 1)
# t.break3 <- c(1991, 26)
t.break1 <- 1989
t.break2 <- 1990
t.break3 <- 1991.5
tb1 <- ts(pmax(0, t - t.break1), start = min(t))
tb2 <- ts(pmax(0, t - t.break2), start = min(t))
tb3 <- ts(pmax(0, t - t.break3), start = min(t))

fit.pw <- tslm(dat ~ t + tb1 + tb2 + tb3)
t.new <- t[length(t)] + seq(h)
tb1.new <- tb1[length(tb1)] + seq(h)
tb2.new <- tb2[length(tb2)] + seq(h)
tb3.new <- tb3[length(tb3)] + seq(h)

newdata <- cbind(t=t.new, tb1=tb1.new, tb2=tb2.new, tb3=tb3.new) %>%
  as.data.frame()
fcasts.pw <- forecast(fit.pw, newdata = newdata)

fit.spline <- tslm(dat ~ t + I(t^2) + I(t^3) +
                     I(tb1^3) + I(tb2^3) + I(tb3^3))
fcasts.spl <- forecast(fit.spline, newdata = newdata)

autoplot(dat, size = 1) +
  autolayer(fitted(fit.lin), series = "Linear", size = 1) +
  autolayer(fitted(fit.exp), series = "Exponential", size = 1) +
  autolayer(fitted(fit.pw), series = "Piecewise", size = 1) +
  autolayer(fitted(fit.spline), series = "Cubic Spline", size = 1) +
  autolayer(fcasts.pw, series="Piecewise") +
  autolayer(fcasts.lin, series="Linear", PI=FALSE) +
  autolayer(fcasts.exp, series="Exponential", PI=FALSE) +
  #autolayer(fcasts.spl, series="Cubic Spline", PI=FALSE) + ## way too high
  xlab("Year") + ylab("Winning times in minutes") +
  ggtitle("Mel-Syd, Economy.Class") +
  guides(colour = guide_legend(title = " "))

checkresiduals(fit.pw)
checkresiduals(fit.spline)
```

# Session info

```{r}
## Packages used
pkgs <- c("oddstream")
## Package & session info
devtools::session_info(pkgs)
```

# Sources & related content

- https://otexts.com/fpp2/
- https://github.com/pridiltal/oddstream
