---
title: "Intro to data.table syntax"
subtitle: "<Option subtitle>"
author: "Nicholas Spyrison"
date: "03 May 2022"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r opts_chunk, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Gist

Learning __data.table__ syntax following the intro vignette; https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html

__data.table__ provides unique data.frame/tibble structures, a parrellized and "self-optimizing" interface for perform common __dplyr__/SQL verbs. While reading is a bit dense, it follows the general form of:

```
DT[i, j, by][...]

##   R:                 i                 j        by
## SQL:  where | order by   select | update  group by
```

keep in mind that there is also __dtplyr__ which tries to marry the speed of __data.table__ and the syntax/behavior of the __dplyr/tidyverse__.

# Setup

....

```{r}
library(data.table)

input <- "https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv"
flights <- fread(input)
flights
dim(flights)
```


# Examples 

## Selecting rows

```{r}
## Get all the flights with “JFK” as the origin airport in the month of June.
ans <- flights[origin == "JFK" & month == 6L]
head(ans)

## Sort flights first by column origin in ascending order, and then by dest in descending order:
ans <- flights[order(origin, -dest)]
head(ans)

## How many trips have been made in 2014 from “JFK” airport in the month of June?
## .N is roughly dplyr::n() or length(var)
ans <- flights[origin == "JFK" & month == 6L, .N]
ans
```

## Selecting columns

```{r}
## Select both arr_delay and dep_delay columns the data.frame way.
ans <- flights[, c("arr_delay", "dep_delay")]
head(ans)

## Select columns named in a variable using the .. prefix
select_cols = c("arr_delay", "dep_delay")
flights[ , ..select_cols] ## Think of .. as, one level up.
## Alternatively, use `with=FALSE`: 
#flights[ , select_cols, with = FALSE]
```


## Chaining, add/update

```{r}
## WHERE/filter then brackets into ordering
ans <- flights[origin == "JFK" & month == 6L][order(origin, -dest)]
ans

## Add/update with :=
flights[, tot_delay := dep_delay + arr_delay][,c(4:5, 12)]
```

## Joins

```{r}
set.seed(42)
big <- data.table( 
  id = LETTERS[2:11],
  a = sample(101:105, 10, replace = TRUE),
  b = sample(200:300, 10)
)
small <- data.table(
  id = LETTERS[1:5],
  y = sample(1:5, 5, replace = TRUE),
  z = sample(10:20, 5) 
)

# Right outer join
big[small, on = .(id)]

# Inner join
small[big, on = .(id), nomatch = NULL]

# Anti-join 
# (return rows of big with no match in small)
big[!small, on = .(id)]
```



# Session info

```{r}
## Packages used
pkgs <- c("data.table")
## Package & session info
devtools::session_info(pkgs)
```

# Sources & related content

- https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html
- https://medium.com/analytics-vidhya/r-data-table-joins-48f00b46ce29
- https://dtplyr.tidyverse.org/