---
title: "ggenealogy Vignette"
subtitle: "And academic genealogy of mathematicians"
author: "Nicholas Spyrison"
date: "09 Jun 2022"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r opts_chunk, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Gist

The site https://www.mathgenealogy.org/, contains over 276,000 observations of Mathematics PhD grads and their supervisors. This is effectively a geneology of mathematical supervision (which should have some sizable effect on thinking, topics, and reading). The R package __ggenealogy__ contains an example dataset from this source and facilitates the consumption and ploting of this type of data.

Given that my thesis was just certified I want to try to see if I can trace up the mathematical genealogy tree to visualize my thought-leading predecessors.


# Setup

```{r}
library(ggenealogy)
library(ggplot2)
library(magrittr)

data("statGeneal", package = "ggenealogy")
df <- statGeneal %>%
  #dplyr::filter(parent != "") %>%
  tibble::as_tibble()
print(df, n=3)
hist(df$gradYear)
```

Ok, about 8k observations where "all the parent-child relationships where both parent and child received an advanced degree of statistics as of June 6, 2015." This may or may-not contain the need people I am looking for.

Note that grad year:

- Is in the range [1864, 2015].
- Median is 5 greater than mean (left skew)


# Where in the world is ...?

Through trial and error I know that Di Cook is not in the data. The original paper does have Thomas Lumley, another professor of interest. But perhaps first I will manual look up Cook's genealogy.

Di, Di's supersivor, and "grand-supervisor" are not in the list, may have to go to plan B, looking at Thomas Lumley. After looking at both parents and children, I know that Thomas has 1 child in the data; Petra Buzkova. From the paper, we can see that the oldest predescor is David Cox.

```{r}
lumley_p <- grepl("Lumley", df$parent, fixed = TRUE)
sum(lumley_p)
df[lumley_p, ]

## Prep the network info, more on this in `As network layout (iGraph)`.
ig <- dfToIG(df)
```

# Finding a path

Let's grab the paths while we are on the topic of names. Actually, if we go all the way to Buzkova, this is the example case in the paper.

```{r}
pathCB <- getPath("David Cox", "Petra Buzkova", ig, df,
                  "gradYear", isDirected = FALSE)
plotPath(pathCB, df, "gradYear", fontFace = 4) +
  xlab("Graduation Year") +
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 10)) +
  scale_x_continuous(expand = c(0.1, 0.2))
```

Good, we have a start. We will want to find a way to traverse the hierarchy to find all of the ancestors without filling in the cousin nodes (or more preferably faintly filling them in). As an example poster, see https://www.mathgenealogy.org/posters/raich.pdf

# Making trees

```{r}
l <- plotAncDes("David Cox", df, mAnc = 1, mDes = 6, vCol = "blue") +
  labs(subtitle = "Interesting, but too many \n  cousins of Thomas Lumley")
r <- plotAncDes("Thomas Lumley", df, mAnc = 6, mDes = 1, vCol = "blue") +
 labs(subtitle =  "Not very interesting, \n  nb only 1:1 relationships")

library(patchwork)
l + r
```

```{r}
plotPathOnAll(pathCB, df, ig, "gradYear",
              bin = 200, nodeSize = 1, pathNodeSize = 2.5,
              nodeCol = "darkgray", edgeCol = "lightgray", 
              animate = TRUE) ## plotly static interaction not animated.
```


# As network layouts (iGraph)



```{r}
ig <- dfToIG(df)
class(ig)
ig
getBasicStatistics(ig)
plot(ig)

```


# Session info

```{r}
## Packages used
pkgs <- c("ggenealogy", "ggplot2")
## Package & session info
devtools::session_info(pkgs)
```

# Sources & related content

- https://www.jstatsoft.org/article/view/v089i13
- https://www.mathgenealogy.org/index.php
