---
title: "Minst and convolutional nueral networks"
subtitle: "Using keras and tensorflow (GPU)"
author: "Nicholas Spyrison"
date: "11 Apr 2022"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r opts_chunk, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  error   = FALSE
)
## Following source and setup:
if(F){
  browseURL("https://rstudio-pubs-static.s3.amazonaws.com/288923_11f4a96b94a74590af0ca460b6ce7648.html#/3")
  #installing 'devtools' package for installing Packages from github
  install.packages('devtools')
  devtools::install_github("rstudio/keras") 
  #installing keras 
  #It will first install tensorflow then keras.
  #The above code will install the keras library from the GitHub repository.
  #Installing Keras for R
  install.packages("rstudio/keras")
  #loading keras in R 
  library(keras)
  install_keras()
  #The R interface to Keras uses TensorFlow as it’s underlying computation engine.
  tensorflow::install_tensorflow(gpu = T)
}
```

# Gist

This is looking the MINST dataset, a series of low resolution hand-darwn digits. It is a standard classification set for image analysis where each pixel can be treated as a variable in grey scale [0, 1]. each image is 28x28 or 784 variables, there are 60000 training obs, and 10000 test obs. More advances takes may aggregate over the surround 9 pixels for example to reduce the size of the variables. I am also excited to use `tensorflow`, I think this should be using GPU computing for parallelization and hopefully near linear performance increase.

Specifically here we will be creating a sequentially Convolutional Nueral Network (CNN) and use that to classify the digits. We have previous installed Python 3.10, keras, and tensorflow with GPU (NVIDIA GPU required)


# Setup and data

```{r}
library(keras)
#Getting started with Keras for R
#The core data structure of Keras is a model, a way to organize layers. 
#The simplest type of model is the Sequential model, a linear stack of layers. 
#For more complex architectures, you should use the Keras functional API, 
#which allows to build arbitrary graphs of layers.

## Data setup -----
#loading the keras inbuilt mnist dataset
data<-dataset_mnist()
str(data)
siz <- object.size(data)
print(siz, units="Mb")

#Training Data
train_x <- data$train$x
train_y <- data$train$y
#Test Data
test_x <- data$test$x
test_y <- data$test$y

#converting a 2D array into a 1D array for feeding
#into the MLP and normalising the matrix
train_x <- array(as.numeric(train_x), dim = c(dim(train_x)[[1]], 784))
test_x  <- array(as.numeric(test_x),  dim = c(dim(test_x)[[1]],  784))
#scale to [0, 1]
train_x <- train_x / 255
test_x  <- test_x  / 255
#target to classification, of 10 digits
train_y <- to_categorical(train_y, 10)
test_y  <- to_categorical(test_y, 10)
```

# Convolutional nueral network structure
```{r}
#Now defining a keras MLP sequential model containing a linear stack of layers
#defining the model with 1 input layer[256 neurons], 1 hidden layer[128 neurons] 
#with dropout rate 0.4 and 1 output layer[10 neurons]
#i.e number of digits from 0 to 9

model <- keras_model_sequential() %>%
  #Input layer-256 units
  #Add a densely-connected NN layer to an output
  layer_dense(units = 256, activation = "relu",
              input_shape = c(784)) %>%
  #dropout layer to prevent Overfitting
  layer_dropout(rate = 0.4) %>%
  #Hidden Layer-128 units
  #Apply an activation function to an output.
  #Relu can only be used for Hidden layers
  layer_dense(units = 128, activation = "relu") %>%
  #dropout layer to prevent Overfitting
  layer_dropout(rate = 0.4) %>%
  #output layer; 10 digits
  layer_dense(units = 10, activation = "softmax")
#softmax activation for Output layer which computes the probabilities for the classes
#Compiling The Model
#Model's summary-showing its architecture
summary(model)
```

# Compile, train, & performance
```{r}
#Compiling the Model and Optimizing the model
#Configure a Keras model for training using compile()
model %>%
  compile(loss ="categorical_crossentropy",
          optimizer = "adam",
          metrics = c("accuracy"))

##Training the Model
#Now let's train the model on the training dataset  
#epochs = No of iterations on a dataset.
#batchsize = Number of samples per gradient update.
tictoc::tic("CNN fit")
history <- model %>% fit(
  train_x, train_y, epochs = 10, batch_size = 128,
  callbacks = callback_tensorboard(log_dir = "logs/run_b"),
  validation_split = 0.2) 
tictoc::toc()
#train on 80% of train set and will evaluate 
#model's metrics such as loss and accuracy on leftover data
#after training --model gives

summary(history)
history$params
history$metrics #gives loss and accuracy metric for each epoch
```

Ok, that looks like we need a local instance of Python 3.7, will have to come back to this another time. For now we'll follow though with the example

# Ploting model performance by epoch

```{r}
#plotting Model - epoch vs acc and Loss
plot(history, labels = T)
which.min(history$metrics$acc)
#Accuracy least for 1st epoch and highest for last epoch-10
plot(x = history$metrics$acc,y = history$metrics$loss,
     pch = 19, col = 'red', type = 'b',
     ylab = "Error on trining Data", xlab = "Accuracy on Training Data")
title("Plot of accuracy vs Loss")
legend("topright", c("Epochs"), col = "red", pch = 19)
#Now Finally evaluating the Model's Performance
#Evaluating model on the Test dataset
score <- model %>%
  evaluate(test_x, test_y, batch_size = 128)

score
```

# Session info

```{r}
## packages used
pkgs <- c(
  "keras"
)

## package & session info
devtools::session_info(pkgs)
```

# Related content

- https://rstudio-pubs-static.s3.amazonaws.com/288923_11f4a96b94a74590af0ca460b6ce7648.html#/3
- https://github.com/anishsingh20/Deep-Learning-in-R-using-Keras-and-Tensorflow-

